---
title: "Workplace Zones Analysis"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

# Introduction

This notebook outlines the analysis of the Irish Central Statistics Office Workplace Zones dataset, and in particular compares it with the Valuation Office dataset.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = FALSE, warning = FALSE, message = FALSE, cache = TRUE)
library(tidyverse)
library(tidymodels)
library(lubridate)
library(magrittr)
library(scales)
library(sf)
library(summarytools)
library(units)

theme_set(theme_light())
```

Load the data. Data obtained from [this](https://www.cso.ie/en/census/census2016reports/workplacezonesand1kmpopulationgrids/) link.

```{r load-data}
wz <- readxl::read_xlsx("data/wz_2016.xlsx")
wz_glossary <- readxl::read_xlsx("data/wz_glossary.xlsx")

wz %<>% 
  select(-93)

wz_glossary %<>% select(name = `FIELD NAME WITHIN TABLES`, description = `DESCRIPTION OF FIELD`) %>% 
  filter(!is.na(name))

wz_glossary %>% 
  write_csv("op/wz_glossary.csv")

wz %>% summarytools::dfSummary() %>%  view()

wz %>% 
  pivot_longer(cols = starts_with("T"),
               names_to = "Variable") %>% 
  write_csv("op/wz_features_long.csv")
```

The next step is to build descriptive proportions from the data. This is necessary so that comparisons can be made across WZs with differing total numbers of people.

```{r build-predictors}

prop_workers <- wz %>% 
  transmute(T11_C1/T11_T)

prop_0_14 <- wz %>% 
  transmute(T1_A1/T1_T)

prop_15_24 <- wz %>% 
  transmute(T1_A2/T1_T)

prop_25_34 <- wz %>% 
  transmute(T1_A3/T1_T)


prop_35_44 <- wz %>% 
  transmute(T1_A4/T1_T)

prop_45_54 <- wz %>% 
  transmute(T1_A5/T1_T)

prop_55_64 <- wz %>% 
  transmute(T1_A6/T1_T)

prop_gt65 <- wz %>% 
  transmute(T1_A7/T1_T)

prop_mot_foot <- wz %>% 
  transmute(T2_M1 / T2_T)

prop_mot_bike <- wz %>% 
  transmute(T2_M2 / T2_T)

prop_mot_bus <- wz %>% 
  transmute(T2_M3 / T2_T)  

prop_mot_rail <- wz %>% 
  transmute(T2_M4 / T2_T)

prop_mot_motorcycle <- wz %>% 
  transmute(T2_M5 / T2_T)

prop_mot_car_driver <- wz %>% 
  transmute(T2_M6 / T2_T)

prop_mot_car_passenger <- wz %>% 
  transmute(T2_M7 / T2_T)

prop_mot_van <- wz %>% 
  transmute(T2_M8 / T2_T)

prop_lh_630 <- wz %>% 
  transmute(T3_T1 / T3_T)

prop_lh_700 <- wz %>% 
  transmute(T3_T2 / T3_T)

prop_lh_730 <- wz %>% 
  transmute(T3_T3 / T3_T)

prop_lh_800 <- wz %>% 
  transmute(T3_T4 / T3_T)

prop_lh_830 <- wz %>% 
  transmute(T3_T5 / T3_T)

prop_lh_900 <- wz %>% 
  transmute(T3_T6 / T3_T)

prop_lh_930 <- wz %>% 
  transmute(T3_T7 / T3_T)

prop_jt_lt15 <- wz %>% 
  transmute(T4_D1 / T4_T)

prop_jt_lt30 <- wz %>% 
  transmute(T4_D2 / T4_T)

prop_jt_lt45 <- wz %>% 
  transmute(T4_D3 / T4_T)

prop_jt_lt60 <- wz %>% 
  transmute(T4_D4 / T4_T)

prop_jt_lt90 <- wz %>% 
  transmute(T4_D5 / T4_T)

prop_jt_gt90 <- wz %>% 
  transmute(T4_D6 / T4_T)

prop_occ_senior <- wz %>% 
  transmute(T5_O1 / T5_T)

prop_occ_prof <- wz %>% 
  transmute(T5_O2 / T5_T)

prop_occ_tech <- wz %>% 
  transmute(T5_O3 / T5_T)

prop_occ_admin <- wz %>% 
  transmute(T5_O4 / T5_T)

prop_occ_service <- wz %>% 
  transmute(T5_O5 / T5_T)

prop_occ_caring <- wz %>% 
  transmute(T5_O6 / T5_T)

prop_occ_retail <- wz %>% 
  transmute(T5_O7 / T5_T)

prop_occ_factory <- wz %>% 
  transmute(T5_O8 / T5_T)

prop_occ_elementary <- wz %>% 
  transmute(T5_O9 / T5_T)

prop_employers <- wz %>% 
  transmute(T7_SG1 / T7_T)

prop_higher_prof <- wz %>% 
  transmute(T7_SG2 / T7_T)

prop_lower_prof <- wz %>% 
  transmute(T7_SG3 / T7_T)

prop_non_manual <- wz %>% 
  transmute(T7_SG4 / T7_T)

prop_manual_skiled <- wz %>% 
  transmute(T7_SG5 / T7_T)

prop_semi_skilled <- wz %>% 
  transmute(T7_SG6 / T7_T)

prop_unskilled <- wz %>% 
  transmute(T7_SG7 / T7_T)

prop_own_account <- wz %>% 
  transmute(T7_SG8 / T7_T)

prop_farmers <- wz %>% 
  transmute(T7_SG9 / T7_T)

prop_agri_workers <- wz %>% 
  transmute(T7_SG10 / T7_T)

prop_ind_agri <- wz %>% 
  transmute(T8_I1 / T8_T)

prop_ind_manufacturing <- wz %>% 
  transmute(T8_I2 / T8_T)

prop_ind_construction <- wz %>% 
  transmute(T8_I1 / T8_T)

prop_ind_agri <- wz %>% 
  transmute(T8_I3 / T8_T)

prop_ind_retail <- wz %>% 
  transmute(T8_I4 / T8_T)

prop_ind_info_finance <- wz %>% 
  transmute(T8_I5 / T8_T)

prop_ind_public <- wz %>% 
  transmute(T8_I6 / T8_T)

prop_ind_edu_health <- wz %>% 
  transmute(T8_I7 / T8_T)

prop_ind_service <- wz %>% 
  transmute(T8_I8 / T8_T)

prop_nat_ireland <- wz %>% 
  transmute(T9_N1 / T9_T)

prop_nat_uk <- wz %>% 
  transmute(T9_N2 / T9_T)

prop_nat_poland <- wz %>% 
  transmute(T9_N3 / T9_T)

prop_nat_other_eu <- wz %>% 
  transmute(T9_N4 / T9_T)

prop_nat_row <- wz %>% 
  transmute(T9_N5 / T9_T)

prop_loe_2ary <- wz %>% 
  transmute(T10_E1 / T10_T)

prop_loe_3ary <- wz %>% 
  transmute(T10_E2 / T10_T)

# assemble these features
wz_features <- 
  bind_cols(wz %>% select(GUID),
            prop_workers,
            prop_0_14,
            prop_15_24,
            prop_25_34,
            prop_35_44,
            prop_45_54,
            prop_55_64,
            prop_gt65,
            prop_mot_bike,
            prop_mot_bus,
            prop_mot_car_driver,
            prop_mot_car_passenger,
            prop_mot_foot,
            prop_mot_motorcycle,
            prop_mot_rail,
            prop_mot_van,
            prop_ind_agri,
            prop_ind_construction,
            prop_ind_edu_health,
            prop_ind_info_finance,
            prop_ind_manufacturing,
            prop_ind_public,
            prop_ind_retail,
            prop_ind_service,
            prop_agri_workers,
            prop_employers,
            prop_farmers,
            prop_higher_prof,
            prop_lower_prof,
            prop_manual_skiled,
            prop_non_manual,
            prop_own_account,
            prop_semi_skilled,
            prop_loe_2ary,
            prop_loe_3ary,
            prop_jt_lt15,
            prop_jt_lt30,
            prop_jt_lt45,
            prop_jt_lt60,
            prop_jt_lt90,
            prop_jt_gt90,
            prop_lh_630,
            prop_lh_700,
            prop_lh_730,
            prop_lh_800,
            prop_lh_830,
            prop_lh_900,
            prop_lh_930,
            prop_occ_admin,
            prop_occ_caring,
            prop_occ_elementary,
            prop_occ_factory,
            prop_occ_prof,
            prop_occ_retail,
            prop_occ_senior,
            prop_occ_service,
            prop_occ_tech,
            prop_nat_ireland,
            prop_nat_uk,
            prop_nat_other_eu,
            prop_nat_poland,
            prop_nat_row)

names(wz_features) <- c("GUID",
            "prop_workers",
            "prop_0_14",
            "prop_15_24",
            "prop_25_34",
            "prop_35_44",
            "prop_45_54",
            "prop_55_64",
            "prop_gt65",
            "prop_mot_bike",
            "prop_mot_bus",
            "prop_mot_car_driver",
            "prop_mot_car_passenger",
            "prop_mot_foot",
            "prop_mot_motorcycle",
            "prop_mot_rail",
            "prop_mot_van",
            "prop_ind_agri",
            "prop_ind_construction",
            "prop_ind_edu_health",
            "prop_ind_info_finance",
            "prop_ind_manufacturing",
            "prop_ind_public",
            "prop_ind_retail",
            "prop_ind_service",
            "prop_agri_workers",
            "prop_employers",
            "prop_farmers",
            "prop_higher_prof",
            "prop_lower_prof",
            "prop_manual_skiled",
            "prop_non_manual",
            "prop_own_account",
            "prop_semi_skilled",
            "prop_loe_2ary",
            "prop_loe_3ary",
            "prop_jt_lt15",
            "prop_jt_lt30",
            "prop_jt_lt45",
            "prop_jt_lt60",
            "prop_jt_lt90",
            "prop_jt_gt90",
            "prop_lh_630",
            "prop_lh_700",
            "prop_lh_730",
            "prop_lh_800",
            "prop_lh_830",
            "prop_lh_900",
            "prop_lh_930",
            "prop_occ_admin",
            "prop_occ_caring",
            "prop_occ_elementary",
            "prop_occ_factory",
            "prop_occ_prof",
            "prop_occ_retail",
            "prop_occ_senior",
            "prop_occ_service",
            "prop_occ_tech",
            "prop_nat_ireland",
            "prop_nat_uk",
            "prop_nat_other_eu",
            "prop_nat_poland",
            "prop_nat_row")

wz_features %<>%
  mutate(across(everything(), ~replace_na(.x, 0.0)))

wz_features %>%
  write_csv("op/wz_features.csv")

```

Now it's time to get working on the spatial data. We will load the Workplace Zone geometry file from a shapefile that we downloaded from the CSO website. We will also get the Valuation Office data, and create a spatial file for each property using the coordinates given. We will then perform a spatial join and sum up the total rateable value for all properties lying inside a Workplace Zone. then we can start some analysis.

Note: we prepared an extract of the Valuation Office data in a previous analysis - this will be reused. The analysis can be seen [here](https://github.com/mick-odonnell/valuation_office)

```{r spatial-data}
wz_geom <- st_read("data/Workplace_Zones_ITM.shp")

wz_geom %>% 
  mutate(area = st_area(.$geometry)) %>%
  st_set_geometry(NULL) %>% 
  select(GUID, area) %>% 
  mutate(area = drop_units(area))-> wz_area

wz_geom %>%
  st_transform(4326) %>% 
  st_write("op/wz_4326.shp")

# we prepared an extract of the Valuation Office data in a previous analysis
valoff <- read_csv("../valuation_office/op/ireland_vo_records.csv")

valoff %>%
  select(property = `Property Number`,
         total_value = Valuation,
         lon = `X ITM`,
         lat = `Y ITM`) %>% 
  distinct(property, .keep_all = TRUE) %>%
  filter(!is.na(lat) & !is.na(lon)) %>% 
  st_as_sf(coords = c("lon", "lat"), crs = 2157) -> valoff_sf
  
# now perform the spatial join

valoff_sf %>% 
  st_join(wz_geom %>% select(GUID),
          join = st_intersects) %>% 
  st_set_geometry(NULL) -> valoff_by_wz

valoff_by_wz %>% write_csv("op/valoff_by_wz.csv")

# calculate the total sum of rateable valuation by WZ
valoff_by_wz %>% 
  group_by(GUID) %>% 
  summarise(prop_count = n(),
            average_valuation = mean(total_value),
            total_value = sum(total_value)) -> wz_valoff_summary
```

We now have a dataset with proportional amounts of each variable, for each Workplace Zone. It might be useful to do some clustering analysis on this table to see if there are commonalities among the data.

The first step is to perform some dimension reduction via Principal Components Analysis. The first plot below shows the first two components - they don't do much to explain the average valuation within a Workplace Zone though...!

```{r wz-pca}
# lets have an initial look at how correlated the variables in the wz dataset are..
library(corrr)

wz_features_w_area <- wz_features %>% left_join(wz_area)

wzcorr <- wz_features_w_area %>% 
  dplyr::select(where(is.numeric)) %>%
  correlate() %>%
  rearrange()

wzcorr

# first determine how many variables to include via PCA

set.seed(20210807)

wz_features_w_area %>% 
  mutate_all(list(~na_if(., Inf))) -> wz_cleaned

pca_rec <- recipe(~., data = wz_cleaned) %>%
  update_role(GUID, new_role = "id") %>%
  step_center(all_predictors()) %>% 
  step_scale(all_predictors()) %>% 
  step_pca(all_predictors(), 
           id = 'pca',
           threshold = 0.85,
           role = "predictor")

pca_prep <- prep(pca_rec)

pca_prep

pca_prep %>% 
  tidy(id = "pca", type = "variance") %>% 
  dplyr::filter(terms == "percent variance") %>% 
  ggplot(aes(x = component, y = value)) + 
  geom_col(fill = "#b6dfe2") + 
  ylab("% of total variance")

tidied_pca <- tidy(pca_prep, 3)

tidied_pca %>%
  filter(component %in% paste0("PC", 1:5)) %>%
  mutate(component = fct_inorder(component)) %>%
  ggplot(aes(terms, value, fill = terms)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(component~., ncol = 1) +
  labs(y = NULL) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Principal Components 1 through 5",
       y = "Magnitude",
       x = "Category")

library(tidytext)

tidied_pca %>%
  filter(component %in% paste0("PC", 1)) %>%
  group_by(component) %>%
  ungroup() %>%
  mutate(terms = reorder_within(terms, abs(value), component)) %>%
  ggplot(aes(abs(value),terms,  fill = value > 0)) +
  geom_col() +
  labs(
    x = "Absolute value of contribution",
    y = NULL, fill = "Positive?"
  )

pca_variance <- tidy(pca_prep, 3, type = "variance")

pca_variance %>% 
  filter(terms == "cumulative percent variance") %>% 
  filter(value < 85) %>% dim()
```

It seems that 85% of the variability is explained by the first 20 Principal Components.

```{r wz-clustering}
# lets perform k-means clustering, with 20 clustering iterations
# we want to use the pca results as the reduced dimensionality to perform our clustering

wz_reduced <- pca_prep[["template"]]
wz_reduced$GUID -> wz_guids

wz_reduced %>% select(-GUID) -> wz_values

kclusts <- tibble(k = 1:20) %>%
  mutate(kclust = map(k, ~kmeans(wz_values, .x)),
    tidied = map(kclust, tidy),
    glanced = map(kclust, glance),
    augmented = map(kclust, augment, wz_values)
  )

clusterings <- 
  kclusts %>%
  unnest(cols = c(glanced))

ggplot(clusterings, aes(k, tot.withinss)) +
  geom_line() +
  geom_point()


# hard to see an optmum cluster number. Let's go with nine as it's a good number for visualisation.
kmeans(wz_values, 9) %>% 
  augment(wz_reduced) %>% 
  select(-starts_with('PC')) -> cluster_assignments_kmeans9

cluster_assignments_kmeans9 %>% 
  write_csv("./op/wz_clusters.csv")

wz_geom %>% 
  left_join(cluster_assignments_kmeans9) %>% 
  select(cluster = .cluster)-> wz_clusters_geom

wz_clusters_geom %>% 
  st_transform(4326) -> wz_clusters_wgs84

#quick leaflet map
library(leaflet)

factpal <- colorFactor(topo.colors(5), cluster_assignments_kmeans5$.cluster)

leaflet(wz_clusters_wgs84) %>% 
  addTiles() %>% 
  addPolygons(stroke = FALSE,
              color = ~factpal(cluster))
```

It would be nice to be able to build a heatmap of how each variable maps to each cluster.

```{r cluster_variables}
wz_features %<>% left_join(cluster_assignments_kmeans9) %>% 
  rename(cluster = .cluster)

mean_by_cluster <-
  wz_features %>% 
  group_by(cluster) %>% 
  summarise(across(starts_with("prop"), ~ mean(.x))) %>% 
  select(-1)

col_means <- wz_features %>% 
  summarise(across(starts_with("prop"), ~mean(.x)))

col_sd <- wz_features %>% 
  summarise(across(starts_with("prop"), ~sd(.x)))


z_scores <- scale(mean_by_cluster, center = col_means, scale = col_sd)

library(RColorBrewer)
heatmap(t(z_scores),
        col=brewer.pal(6,'BrBG'),
        breaks=c(-1e10,-2,-1,0,1,2,+1e10),
        xlab='Cluster Number',
        add.expr=abline(h=(0:40)+0.5,v=(0:n.clus)+0.5,col='white')) 

z_scores %>% 
  as_tibble() %>% 
  rownames_to_column(var = "cluster") %>% 
  pivot_longer(cols = starts_with("prop"),
               names_to = "predictor") -> z_scores_long

z_scores_long %>% 
  write_csv("op/cluster_weightings_long.csv")
```


We now have a table with summary information for;

-   number of rateable properties by WZ
-   average value of rateable properties
-   total value of rateable properties
